---
title: "M Carlin: First Deliverable"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
setwd("C:/Users/CarlinML/DACSS-690R")
getwd()
```

## FIRST DATASET - CLEANING
# Social Media and Mental Health
```{r}
folder="dataFiles"
fileName="Social_Media_and_Mental_Health.csv"
fileToRead=file.path(folder,fileName)
fileToRead
SMMH_dirty=read.csv(fileToRead,check.names=F)
View(SMMH_dirty)
```

# Remove leading and trailing spaces
```{r}
SMMH_dirty[,]=sapply(SMMH_dirty[,],trimws)
SMMH_clean=SMMH_dirty[,]
View(SMMH_clean)
str(SMMH_clean)
```

# Fix column names
```{r}
library(magrittr)
pattern='\\s+'
names(SMMH_clean)%>%gsub(pattern,'_',.)
```

# Variable names are too cumbersome; decided to use the following code to rename all variables.
```{r}
library("dplyr")
SMMH_clean <- rename(SMMH_clean, 
  Age = "1. What is your age?",
  Gender = "2. Gender",
  Relationship_Status = "3. Relationship Status",
  Occupation_Status = "4. Occupation Status",
  Org_Affiliation = "5. What type of organizations are you affiliated with?",
  Use_Social_Media = "6. Do you use social media?",
  Social_Media_Platforms = "7. What social media platforms do you commonly use?",
  Avg_Amt_Time = "8. What is the average time you spend on social media every day?",
  Without_Purpose = "9. How often do you find yourself using Social media without a specific purpose?",
  Distracted_By = "10. How often do you get distracted by Social media when you are busy doing something?",
  Feel_Restless = "11. Do you feel restless if you haven't used Social media in a while?",
  Easily_Distracted = "12. On a scale of 1 to 5, how easily distracted are you?",
  Bothered_Worries = "13. On a scale of 1 to 5, how much are you bothered by worries?",
  Difficult_Concentrate = "14. Do you find it difficult to concentrate on things?",
  Compare_Others = "15. On a scale of 1-5, how often do you compare yourself to other successful people through the use of social media?",
  Feel_Comparisons = "16. Following the previous question, how do you feel about these comparisons, generally speaking?",
  Seek_Validation = "17. How often do you look to seek validation from features of social media?",
  Feel_Depressed = "18. How often do you feel depressed or down?",
  Interest_Fluctuate = "19. On a scale of 1 to 5, how frequently does your interest in daily activities fluctuate?",
  Sleep_Issues = "20. On a scale of 1 to 5, how often do you face issues regarding sleep?")
```

# Run frequency tables of categorical variables to see if any recoding is needed
```{r}
table(SMMH_clean$Gender)
SMMH_clean[SMMH_clean$Gender=='NB','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='Non binary','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='Nonbinary','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='unsure','Gender']='NA'
SMMH_clean[SMMH_clean$Gender=='There are others???','Gender']='NA'
table(SMMH_clean$Gender)
```

# Additional freq tables and mispelling correction
```{r}
table(SMMH_clean$Relationship_Status) # no cleaning needed
table(SMMH_clean$Occupation_Status) # no cleaning needed
table(SMMH_clean$Org_Affiliation) # clean up N/As
SMMH_clean$Org_Affiliation <- gsub("Goverment", "Government",
                              gsub("N/A", "NA", 
                              gsub(", N/A", "", SMMH_clean$Org_Affiliation)))
table(SMMH_clean$Org_Affiliation)
```

# Additional freq tables
```{r}
table(SMMH_clean$Use_Social_Media) # no cleaning needed
table(SMMH_clean$Social_Media_Platforms) # may need to clean but would be easier if data was formatted differently
table(SMMH_clean$Avg_Amt_Time) # no cleaning needed
```

# Save cleaned file to new folder
```{r}
folder <- "Second_Deliverable/DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(SMMH_clean,file.path(folder,"SMMH_clean.csv"),row.names = F)

} else {
  write.csv(SMMH_clean,file.path(folder,"SMMH_clean.csv"),row.names = F)}
```

## FIRST DATASET - FORMATTING
```{r}

```



# SECOND DATASET - Clerkship grades; need to skip first line in excel file
```{r}
library(readxl)
folder="dataFiles"
fileName="UMassChan_ClerkshipGrades.xlsx"
fileToRead=file.path(folder,fileName)
fileToRead
ClerkshipGrades_dirty=read_xlsx(fileToRead, skip=1)
View(ClerkshipGrades_dirty)
```

# Remove leading and trailing spaces
```{r}
ClerkshipGrades_dirty[,]=sapply(ClerkshipGrades_dirty[,],trimws)
ClerkshipGrades_clean=ClerkshipGrades_dirty[,]
View(ClerkshipGrades_clean)
str(ClerkshipGrades_clean)
```

# Remove dashes from two variable names
```{r}
ClerkshipGrades_clean <- rename(ClerkshipGrades_clean, 
  NBME_Letter = `NBME-Written_Letter`,
  NBME_Numeric = `NBME-Written_Numeric`)
```

# Run frequency tables of categorical variables to see if any recoding is needed
```{r}
table(ClerkshipGrades_clean$Final_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(Final_Letter2 = recode(Final_Letter,
                                  "HH" = "High Honors",
                                  "H" = "Honors",
                                  "P" = "Pass",
                                  "S" = "Satisfactory",
                                  "F" = "Fail",
                                  "I" = "Incomplete"))
table(ClerkshipGrades_clean$Final_Letter2)

table(ClerkshipGrades_clean$SPE_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(SPE_Letter2 = recode(SPE_Letter,
                                  "HONORS" = "Honors"))
table(ClerkshipGrades_clean$SPE_Letter2)

table(ClerkshipGrades_clean$NBME_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(NBME_Letter2 = recode(NBME_Letter,
                                  "HIGH HONORS" = "High Honors",
                                  "HONORS" = "Honors",
                                  "PASS" = "Pass"))
table(ClerkshipGrades_clean$NBME_Letter2)

table(ClerkshipGrades_clean$OSCE_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(OSCE_Letter2 = recode(OSCE_Letter2,
                                  "HIGH HONORS" = "High Honors",
                                  "HONORS" = "Honors",
                                  "PASS" = "Pass"))
table(ClerkshipGrades_clean$OSCE_Letter2)
```

# Save cleaned file to new folder
```{r}
folder <- "Second_Deliverable/DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(SMMH_clean,file.path(folder,"ClerkshipGrades_clean.csv"))

} else {
  write.csv(ClerkshipGrades_clean,file.path(folder,"ClerkshipGrades_clean.csv"))}
```


# THIRD DATASET - subset of crime data
```{r}
#install.packages("jsonlite")
library(jsonlite)
endPoint="https://data.lacity.org/resource/2nrs-mtv8.json"
LA_Crime_data = fromJSON(endPoint)
View(LA_Crime_data)
write.csv(LA_Crime_data, "../dataFiles/Crime_Data_subset.csv")
```

