---
title: "M Carlin: First Deliverable"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
setwd("C:/Users/CarlinML/DACSS-690R")
getwd()
```

## FIRST DATASET - CLEANING
# Social Media and Mental Health
```{r}
folder="dataFiles"
fileName="Social_Media_and_Mental_Health.csv"
fileToRead=file.path(folder,fileName)
fileToRead
SMMH_dirty=read.csv(fileToRead,check.names=F)
SMMH_dirty
```

# Remove leading and trailing spaces
```{r}
SMMH_dirty[,]=sapply(SMMH_dirty[,],trimws)
SMMH_clean=SMMH_dirty[,]
View(SMMH_clean)
str(SMMH_clean)
```

# Fix column names (variable names were too cumbersome when I tried replacing spaces with underscores, so I decided to use the following code to rename the variables instead)
```{r}
library("dplyr")
SMMH_clean <- rename(SMMH_clean, 
  Age = "1. What is your age?",
  Gender = "2. Gender",
  RelStatus = "3. Relationship Status",
  OccStatus = "4. Occupation Status",
  OrgAffil = "5. What type of organizations are you affiliated with?",
  UseSocialMedia = "6. Do you use social media?",
  Platforms = "7. What social media platforms do you commonly use?",
  AmtTime = "8. What is the average time you spend on social media every day?",
  WoutPurpose = "9. How often do you find yourself using Social media without a specific purpose?",
  Distracted = "10. How often do you get distracted by Social media when you are busy doing something?",
  Restless = "11. Do you feel restless if you haven't used Social media in a while?",
  EasilyDist = "12. On a scale of 1 to 5, how easily distracted are you?",
  Bothered = "13. On a scale of 1 to 5, how much are you bothered by worries?",
  DiffConcen = "14. Do you find it difficult to concentrate on things?",
  Compare = "15. On a scale of 1-5, how often do you compare yourself to other successful people through the use of social media?",
  Comparisons = "16. Following the previous question, how do you feel about these comparisons, generally speaking?",
  Validation = "17. How often do you look to seek validation from features of social media?",
  Depressed = "18. How often do you feel depressed or down?",
  Interest = "19. On a scale of 1 to 5, how frequently does your interest in daily activities fluctuate?",
  Sleep = "20. On a scale of 1 to 5, how often do you face issues regarding sleep?")

```

# Run frequency tables of categorical variables to see if any recoding is needed
```{r}
table(SMMH_clean$Gender)
SMMH_clean[SMMH_clean$Gender=='NB','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='Non binary','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='Nonbinary','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='unsure','Gender']='NA'
SMMH_clean[SMMH_clean$Gender=='There are others???','Gender']='NA'
table(SMMH_clean$Gender)
```

# Additional freq tables and mispelling correction
```{r}
table(SMMH_clean$RelStatus) # no cleaning needed
table(SMMH_clean$OccStatus) # no cleaning needed
table(SMMH_clean$OrgAffil) # clean up N/As
SMMH_clean$OrgAffil <- gsub("Goverment", "Government",
                              gsub("N/A", "NA", 
                              gsub(", N/A", "", SMMH_clean$OrgAffil)))
table(SMMH_clean$OrgAffil)
```

# Additional freq tables
```{r}
table(SMMH_clean$UseSocialMedia) # no cleaning needed
table(SMMH_clean$Platforms) # may need to clean but would be easier if data was formatted differently
table(SMMH_clean$AmtTime) # no cleaning needed
```

# Save cleaned file to new folder
```{r}
folder <- "Second_Deliverable/DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(SMMH_clean,file.path(folder,"SMMH_clean.csv"),row.names = F)

} else {
  write.csv(SMMH_clean,file.path(folder,"SMMH_clean.csv"),row.names = F)}
```

## FIRST DATASET - FORMATTING
```{r}
linkSMMH_clean='https://github.com/DACSS-690R/Second_Deliverable/raw/refs/heads/main/DataCleanAndFormatted/SMMH_clean.csv'
SMMH_clean=read.csv(linkSMMH_clean)
str(SMMH_clean)
```

# Nominal variables - create new columns that are factors
```{r}
SMMH_clean$Gender_label <- as.factor(SMMH_clean$Gender)
SMMH_clean$RelStatus_label <- as.factor(SMMH_clean$RelStatus)
SMMH_clean$OccStatus_label <- as.factor(SMMH_clean$OccStatus)
SMMH_clean$OrgAffil_label <- as.factor(SMMH_clean$OrgAffil)
SMMH_clean$UseSocialMedia_label <- as.factor(SMMH_clean$UseSocialMedia)
str(SMMH_clean)
```

# Ordinal 'amount of time' variable
```{r}
# map for replacement: 1 the lowest # hours / 6 the highest # hours
map_AmtTime <- c("Less than an Hour"=1, "Between 1 and 2 hours"=2,"Between 2 and 3 hours"=3, "Between 3 and 4 hours"=4, "Between 4 and 5 hours"=5, "More than 5 hours"=6)

SMMH_clean$AmtTime_int=map_AmtTime[SMMH_clean$AmtTime]

SMMH_clean$AmtTime_label <- factor(SMMH_clean$AmtTime_int, levels = seq(1,6),labels = c('1_Less than an Hour','2_Between 1 and 2 hours','3_Between 2 and 3 hours','4_Between 3 and 4 hours','5_Between 4 and 5 hours', '6_More than 5 hours'),ordered = TRUE)
```

# Ordinal likert scale items
```{r}
theInts=seq(1,5) 
theLabels=c('1_Strongly Disagree','2_Disagree','3_Neutral','4_Agree','5_Strongly Agree') 

FormatOrdinal=function(col) factor(col,
                                   levels = theInts,
                                   labels = theLabels,
                                   ordered = TRUE)

names <- c(10:21)
SMMH_clean[names]=lapply(SMMH_clean[names],FormatOrdinal)

str(SMMH_clean)
```

# Save as RDS
```{r}
folder = "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  saveRDS(SMMH_clean,file.path(folder,"SMMH_formatted.RDS"))

} else {
  saveRDS(SMMH_clean,file.path(folder,"SMMH_formatted.RDS"))

}
```


# SECOND DATASET - Clerkship grades; need to skip first line in excel file
```{r}
library(readxl)
folder="dataFiles"
fileName="UMassChan_ClerkshipGrades.xlsx"
fileToRead=file.path(folder,fileName)
fileToRead
ClerkshipGrades_dirty=read_xlsx(fileToRead, skip=1)
ClerkshipGrades_dirty
```

# Remove leading and trailing spaces
```{r}
ClerkshipGrades_dirty[,]=sapply(ClerkshipGrades_dirty[,],trimws)
ClerkshipGrades_clean=ClerkshipGrades_dirty[,]
View(ClerkshipGrades_clean)
str(ClerkshipGrades_clean)
```

# Remove dashes from two variable names
```{r}
ClerkshipGrades_clean <- rename(ClerkshipGrades_clean, 
  NBME_Letter = `NBME-Written_Letter`,
  NBME_Numeric = `NBME-Written_Numeric`)
```

# Run frequency tables of categorical variables to see if any recoding is needed
```{r}
table(ClerkshipGrades_clean$Subject) # no cleaning needed
table(ClerkshipGrades_clean$Catalog) # no cleaning needed 
table(ClerkshipGrades_clean$Session) # no cleaning needed
table(ClerkshipGrades_clean$Location) # no cleaning needed
table(ClerkshipGrades_clean$Final_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(Final_Letter2 = recode(Final_Letter,
                                  "HH" = "High Honors",
                                  "H" = "Honors",
                                  "P" = "Pass",
                                  "S" = "Satisfactory",
                                  "F" = "Fail",
                                  "I" = "Incomplete"))
table(ClerkshipGrades_clean$Final_Letter2)

table(ClerkshipGrades_clean$SPE_Letter) 
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(SPE_Letter2 = recode(SPE_Letter,
                                  "HONORS" = "Honors"))
table(ClerkshipGrades_clean$SPE_Letter2)

table(ClerkshipGrades_clean$NBME_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(NBME_Letter2 = recode(NBME_Letter,
                                  "HIGH HONORS" = "High Honors",
                                  "HONORS" = "Honors",
                                  "PASS" = "Pass"))
table(ClerkshipGrades_clean$NBME_Letter2)

table(ClerkshipGrades_clean$OSCE_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(OSCE_Letter2 = recode(OSCE_Letter,
                                  "HIGH HONORS" = "High Honors",
                                  "HONORS" = "Honors",
                                  "PASS" = "Pass"))
table(ClerkshipGrades_clean$OSCE_Letter2)
```

# Cleaning numeric variables
```{r}
ClerkshipGrades_clean[!complete.cases(ClerkshipGrades_clean),] 
ClerkshipGrades_clean[1,]
colSums(is.na(apply(ClerkshipGrades_clean[,c(10,12,14,16)],2, as.numeric)))

detectWrongNA= function(col){col[grep("[^\\d+\\.*\\d*]", col, perl=T,invert = F)]}
badSymbolNum=sapply(ClerkshipGrades_clean[, c('Final_Numeric','SPE_Numeric','NBME_Numeric', 'OSCE_Numeric')],detectWrongNA)
badSymbolNum_unlist=unlist(badSymbolNum)
badSymbolNum_vector=unique(badSymbolNum_unlist)
badSymbolNum_vector

ClerkshipGrades_clean[, c('Final_Numeric','SPE_Numeric','NBME_Numeric', 'OSCE_Numeric')]=lapply(ClerkshipGrades_clean[, c('Final_Numeric','SPE_Numeric','NBME_Numeric', 'OSCE_Numeric')],function(col) ifelse((col %in% badSymbolNum_vector), NA, col))

ClerkshipGrades_clean
```

# Save cleaned file to new folder
```{r}
folder <- "Second_Deliverable/DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(SMMH_clean,file.path(folder,"ClerkshipGrades_clean.csv"))

} else {
  write.csv(ClerkshipGrades_clean,file.path(folder,"ClerkshipGrades_clean.csv"))}
```

# SECOND DATASET - FORMATTING
```{r}
linkClerkshipGrades_clean='https://github.com/DACSS-690R/Second_Deliverable/raw/refs/heads/main/DataCleanAndFormatted/ClerkshipGrades_clean.csv'
ClerkshipGrades_clean=read.csv(linkClerkshipGrades_clean)
str(ClerkshipGrades_clean)
```

# Change all Letter Grade columns to uppercase, and conver to ordered factors
```{r}

ClerkshipGrades_clean[,16:19] <- lapply(ClerkshipGrades_clean[,16:19],toupper)


Likert_cols <- c(16:19)
ClerkshipGrades_clean[,Likert_cols] <- lapply(ClerkshipGrades_clean[,Likert_cols] , factor, ordered = TRUE, levels = c("FAIL", "INCOMPLETE", "SATISFACTORY", "PASS", "HONORS", "HIGH HONORS"))
str(ClerkshipGrades_clean)
```

# Save as RDS
```{r}
folder = "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  saveRDS(ClerkshipGrades_clean,file.path(folder,"
                               ClerkshipGrades_formatted.RDS"))

} else {
  saveRDS(ClerkshipGrades_clean,file.path(folder,"ClerkshipGrades_formatted.RDS"))

}
```


# THIRD DATASET - subset of crime data
```{r}
#install.packages("jsonlite")
library(jsonlite)
endPoint="https://data.lacity.org/resource/2nrs-mtv8.json"
LA_Crime_data = fromJSON(endPoint)
View(LA_Crime_data)
write.csv(LA_Crime_data, "../dataFiles/Crime_Data_subset.csv")
```
