---
title: 'M Carlin: Second Deliverable'
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r}
setwd("C:/Users/CarlinML/DACSS-690R/Second_Deliverable")
getwd()
```

## FIRST DATASET - CLEANING

# Social Media and Mental Health

```{r}
SMMH_dirty <- read.csv("https://github.com/DACSS-690R/First_Deliverable/raw/refs/heads/main/dataFiles/Social_Media_and_Mental_Health.csv",check.names=F)
SMMH_dirty
```

# Remove leading and trailing spaces

```{r}
SMMH_dirty[,]=sapply(SMMH_dirty[,],trimws)
SMMH_clean=SMMH_dirty[,]
View(SMMH_clean)
str(SMMH_clean)
```

# Fix column names (variable names were too cumbersome when I tried replacing spaces with underscores, so I decided to use the following code to rename the variables instead)

```{r}
library(dplyr)
SMMH_clean <- rename(SMMH_clean, 
  Age = "1. What is your age?",
  Gender = "2. Gender",
  RelStatus = "3. Relationship Status",
  OccStatus = "4. Occupation Status",
  OrgAffil = "5. What type of organizations are you affiliated with?",
  UseSocialMedia = "6. Do you use social media?",
  Platforms = "7. What social media platforms do you commonly use?",
  AmtTime = "8. What is the average time you spend on social media every day?",
  WoutPurpose = "9. How often do you find yourself using Social media without a specific purpose?",
  Distracted = "10. How often do you get distracted by Social media when you are busy doing something?",
  Restless = "11. Do you feel restless if you haven't used Social media in a while?",
  EasilyDist = "12. On a scale of 1 to 5, how easily distracted are you?",
  Bothered = "13. On a scale of 1 to 5, how much are you bothered by worries?",
  DiffConcen = "14. Do you find it difficult to concentrate on things?",
  Compare = "15. On a scale of 1-5, how often do you compare yourself to other successful people through the use of social media?",
  Comparisons = "16. Following the previous question, how do you feel about these comparisons, generally speaking?",
  Validation = "17. How often do you look to seek validation from features of social media?",
  Depressed = "18. How often do you feel depressed or down?",
  Interest = "19. On a scale of 1 to 5, how frequently does your interest in daily activities fluctuate?",
  Sleep = "20. On a scale of 1 to 5, how often do you face issues regarding sleep?")

```

# Format Timestamp variable to be a date column

```{r}
SMMH_clean$Timestamp=strptime(SMMH_clean$Timestamp, "%m/%d/%Y %H:%M:%S")
str(SMMH_clean)
```

# Run frequency tables of categorical variables to see if any recoding is needed

```{r}
table(SMMH_clean$Gender)
SMMH_clean[SMMH_clean$Gender=='NB','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='Non binary','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='Nonbinary','Gender']='Non-binary'
SMMH_clean[SMMH_clean$Gender=='unsure','Gender']='NA'
SMMH_clean[SMMH_clean$Gender=='There are others???','Gender']='NA'
table(SMMH_clean$Gender)
```

# Additional freq tables and mispelling correction

```{r}
table(SMMH_clean$RelStatus) # no cleaning needed
table(SMMH_clean$OccStatus) # no cleaning needed
table(SMMH_clean$OrgAffil) # clean up N/As
SMMH_clean$OrgAffil <- gsub("Goverment", "Government",
                              gsub("N/A", "NA", 
                              gsub(", N/A", "", SMMH_clean$OrgAffil)))
table(SMMH_clean$OrgAffil)
```

# Additional freq tables

```{r}
table(SMMH_clean$UseSocialMedia) # no cleaning needed
table(SMMH_clean$Platforms) # may need to clean but would be easier if data was formatted differently
table(SMMH_clean$AmtTime) # no cleaning needed
```

# Save cleaned file to new folder

```{r}
folder <- "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(SMMH_clean,file.path(folder,"SMMH_clean.csv"),row.names = F)

} else {
  write.csv(SMMH_clean,file.path(folder,"SMMH_clean.csv"),row.names = F)}
```

## FIRST DATASET - FORMATTING

```{r}
linkSMMH_clean='https://github.com/DACSS-690R/Second_Deliverable/raw/refs/heads/main/DataCleanAndFormatted/SMMH_clean.csv'
SMMH_clean=read.csv(linkSMMH_clean)
str(SMMH_clean)
```

# Nominal variables - create new columns that are factors

```{r}
SMMH_clean$Gender_label <- as.factor(SMMH_clean$Gender)
SMMH_clean$RelStatus_label <- as.factor(SMMH_clean$RelStatus)
SMMH_clean$OccStatus_label <- as.factor(SMMH_clean$OccStatus)
SMMH_clean$OrgAffil_label <- as.factor(SMMH_clean$OrgAffil)
SMMH_clean$UseSocialMedia_label <- as.factor(SMMH_clean$UseSocialMedia)
str(SMMH_clean)
```

# Ordinal 'amount of time' variable

```{r}
# map for replacement: 1 the lowest # hours / 6 the highest # hours
map_AmtTime <- c("Less than an Hour"=1, "Between 1 and 2 hours"=2,"Between 2 and 3 hours"=3, "Between 3 and 4 hours"=4, "Between 4 and 5 hours"=5, "More than 5 hours"=6)

SMMH_clean$AmtTime_int=map_AmtTime[SMMH_clean$AmtTime]

SMMH_clean$AmtTime_label <- factor(SMMH_clean$AmtTime_int, levels = seq(1,6),labels = c('1_Less than an Hour','2_Between 1 and 2 hours','3_Between 2 and 3 hours','4_Between 3 and 4 hours','5_Between 4 and 5 hours', '6_More than 5 hours'),ordered = TRUE)
```

# Ordinal likert scale items

```{r}
theInts=seq(1,5) 
theLabels=c('1_Strongly Disagree','2_Disagree','3_Neutral','4_Agree','5_Strongly Agree') 

FormatOrdinal=function(col) factor(col,
                                   levels = theInts,
                                   labels = theLabels,
                                   ordered = TRUE)

names <- c(10:21)
SMMH_clean[names]=lapply(SMMH_clean[names],FormatOrdinal)

str(SMMH_clean)
```

# Save as RDS

```{r}
folder = "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  saveRDS(SMMH_clean,file.path(folder,"SMMH_formatted.RDS"))

} else {
  saveRDS(SMMH_clean,file.path(folder,"SMMH_formatted.RDS"))

}
```

## SECOND DATASET - CLEANING

# Clerkship grades; need to skip first line in excel file

```{r}
library(readxl)

destfile <- path.expand("ClerkshipGrades.xlsx")
url      <- paste0("https://github.com/DACSS-690R/First_Deliverable/raw/refs/heads/main/dataFiles/UMassChan_ClerkshipGrades.xlsx")

download.file(url, destfile = destfile, mode = "wb")
ClerkshipGrades_dirty <- read_xlsx(destfile, skip=1)
ClerkshipGrades_dirty
```

# Remove leading and trailing spaces

```{r}
ClerkshipGrades_dirty[,]=sapply(ClerkshipGrades_dirty[,],trimws)
ClerkshipGrades_clean=ClerkshipGrades_dirty[,]
View(ClerkshipGrades_clean)
str(ClerkshipGrades_clean)
```

# Remove dashes from two variable names

```{r}
ClerkshipGrades_clean <- rename(ClerkshipGrades_clean, 
  NBME_Letter = `NBME-Written_Letter`,
  NBME_Numeric = `NBME-Written_Numeric`)
```

# Run frequency tables of categorical variables to see if any recoding is needed

```{r}
table(ClerkshipGrades_clean$Subject) # no cleaning needed
table(ClerkshipGrades_clean$Catalog) # no cleaning needed 
table(ClerkshipGrades_clean$Session) # no cleaning needed
table(ClerkshipGrades_clean$Location) # no cleaning needed
table(ClerkshipGrades_clean$Final_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(Final_Letter2 = recode(Final_Letter,
                                  "HH" = "High Honors",
                                  "H" = "Honors",
                                  "P" = "Pass",
                                  "S" = "Satisfactory",
                                  "F" = "Fail",
                                  "I" = "Incomplete"))
table(ClerkshipGrades_clean$Final_Letter2)

table(ClerkshipGrades_clean$SPE_Letter) 
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(SPE_Letter2 = recode(SPE_Letter,
                                  "HONORS" = "Honors"))
table(ClerkshipGrades_clean$SPE_Letter2)

table(ClerkshipGrades_clean$NBME_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(NBME_Letter2 = recode(NBME_Letter,
                                  "HIGH HONORS" = "High Honors",
                                  "HONORS" = "Honors",
                                  "PASS" = "Pass"))
table(ClerkshipGrades_clean$NBME_Letter2)

table(ClerkshipGrades_clean$OSCE_Letter)
ClerkshipGrades_clean <- ClerkshipGrades_clean %>%
  mutate(OSCE_Letter2 = recode(OSCE_Letter,
                                  "HIGH HONORS" = "High Honors",
                                  "HONORS" = "Honors",
                                  "PASS" = "Pass"))
table(ClerkshipGrades_clean$OSCE_Letter2)
```

# Cleaning numeric variables

```{r}
ClerkshipGrades_clean[!complete.cases(ClerkshipGrades_clean),] 
ClerkshipGrades_clean[1,]
colSums(is.na(apply(ClerkshipGrades_clean[,c(10,12,14,16)],2, as.numeric)))

detectWrongNA= function(col){col[grep("[^\\d+\\.*\\d*]", col, perl=T,invert = F)]}
badSymbolNum=sapply(ClerkshipGrades_clean[, c('Final_Numeric','SPE_Numeric','NBME_Numeric', 'OSCE_Numeric')],detectWrongNA)
badSymbolNum_unlist=unlist(badSymbolNum)
badSymbolNum_vector=unique(badSymbolNum_unlist)
badSymbolNum_vector

ClerkshipGrades_clean[, c('Final_Numeric','SPE_Numeric','NBME_Numeric', 'OSCE_Numeric')]=lapply(ClerkshipGrades_clean[, c('Final_Numeric','SPE_Numeric','NBME_Numeric', 'OSCE_Numeric')],function(col) ifelse((col %in% badSymbolNum_vector), NA, col))

ClerkshipGrades_clean
```

# Save cleaned file to new folder

```{r}
folder <- "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(SMMH_clean,file.path(folder,"ClerkshipGrades_clean.csv"))

} else {
  write.csv(ClerkshipGrades_clean,file.path(folder,"ClerkshipGrades_clean.csv"))}
```

# SECOND DATASET - FORMATTING

```{r}
linkClerkshipGrades_clean='https://github.com/DACSS-690R/Second_Deliverable/raw/refs/heads/main/DataCleanAndFormatted/ClerkshipGrades_clean.csv'
ClerkshipGrades_clean=read.csv(linkClerkshipGrades_clean)
str(ClerkshipGrades_clean)
```

# Change all Letter Grade columns to uppercase, and convert to ordered factors

```{r}

ClerkshipGrades_clean[,16:19] <- lapply(ClerkshipGrades_clean[,16:19],toupper)


Likert_cols <- c(16:19)
ClerkshipGrades_clean[,Likert_cols] <- lapply(ClerkshipGrades_clean[,Likert_cols] , factor, ordered = TRUE, levels = c("FAIL", "INCOMPLETE", "SATISFACTORY", "PASS", "HONORS", "HIGH HONORS"))
str(ClerkshipGrades_clean)
```

# Save as RDS

```{r}
folder = "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  saveRDS(ClerkshipGrades_clean,file.path(folder,"
                               ClerkshipGrades_formatted.RDS"))

} else {
  saveRDS(ClerkshipGrades_clean,file.path(folder,"ClerkshipGrades_formatted.RDS"))

}
```

## THIRD DATASET - CLEANING

# Subset of crime data

```{r}
Crime_dirty <- read.csv("https://github.com/DACSS-690R/First_Deliverable/raw/refs/heads/main/dataFiles/Crime_Data_from_2020_to_Present_subset.csv",check.names=F)
Crime_dirty
```

# Remove leading/trailing spaces and drop first column

```{r}
Crime_dirty[,]=sapply(Crime_dirty[,],trimws)
Crime_clean=Crime_dirty[,]
Crime_clean=Crime_clean[,2:28]
View(Crime_clean)
str(Crime_clean)
```

```{r}
Crime_clean <- rename(Crime_clean, 
  RecordNumber = "dr_no",
  DateReported = "date_rptd",
  DateOccurred = "date_occ",
  TimeOccurred = "time_occ",
  GeographicAreaID = "area",
  GeographicAreaName = "area_name",
  GeographicSubAreaID = "rpt_dist_no",
  Part1_2 = "part_1_2",
  CrimeCode = "crm_cd",
  CrimeCodeDesc = "crm_cd_desc",
  VictimAge = "vict_age",
  VictimSex = "vict_sex",
  VictimEthnicity = "vict_descent",
  CrimeSceneCode = "premis_cd",
  CrimeSceneDesc = "premis_desc",
  CaseStatus = "status",
  CaseStatusDesc = "status_desc",
  CrimeCode1 = "crm_cd_1",
  CrimeCode2 = "crm_cd_2",
  AddressCrime = "location",
  LatitudeCrime = "lat",
  LongitudeCrime = "lon",
  CriminalCharacteristicCodes = "mocodes",
  WeaponUsedCode = "weapon_used_cd",
  WeaponUsedDesc = "weapon_desc",
  CrossStreet = "cross_street",
  CrimeCode3 = "crm_cd_3")

```

# Format DateReported and DateOccurred variables to be date columns

FORMAT TIME!!! format(strptime(Crime_clean$TimeOccurred, '%H%M'), '%I:%M %p')


```{r}
Crime_clean$DateReported=strptime(Crime_clean$DateReported, "%Y-%m-%dT%H:%M:%S")
Crime_clean$DateOccurred=strptime(Crime_clean$DateOccurred, "%Y-%m-%dT%H:%M:%S")
str(Crime_clean)
```

# Run frequency tables of all variables to see if any cleaning is needed

```{r}
table(Crime_clean$GeographicAreaID) #no cleaning needed
table(Crime_clean$GeographicAreaName) #no cleaning needed
table(Crime_clean$GeographicSubAreaID) #no cleaning needed
table(Crime_clean$Part1_2) #no cleaning needed
table(Crime_clean$CrimeCode) #no cleaning needed
table(Crime_clean$CrimeCodeDesc) #no cleaning needed
table(Crime_clean$VictimeAge) #no cleaning needed
table(Crime_clean$VictimSex) #no cleaning needed
table(Crime_clean$VictimEthnicity) #no cleaning needed
table(Crime_clean$CrimeSceneCode) #no cleaning needed
table(Crime_clean$CrimeSceneDesc) #no cleaning needed
table(Crime_clean$CaseStatus) #no cleaning needed
table(Crime_clean$CaseStatusDesc) #no cleaning needed
table(Crime_clean$CrimeCode1) #no cleaning needed
table(Crime_clean$CrimeCode2) #no cleaning needed
table(Crime_clean$AddressCrime) #no cleaning needed
table(Crime_clean$LatitudeCrime) #no cleaning needed
table(Crime_clean$LongitudeCrime) #no cleaning needed
table(Crime_clean$CriminalCharacteristicCodes) #no cleaning needed
table(Crime_clean$WeaponUsedCode) #no cleaning needed
table(Crime_clean$WeaponUsedDesc) #no cleaning needed
table(Crime_clean$CrossStreet) #no cleaning needed
table(Crime_clean$CrimeCode3) #no cleaning needed
```

# Save cleaned file to new folder

```{r}
folder <- "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  write.csv(Crime_clean,file.path(folder,"Crime_clean.csv"))

} else {
  write.csv(Crime_clean,file.path(folder,"Crime_clean.csv"))}
```

# THIRD DATASET - FORMATTING

# remove extra blank spaces from AddressCrime; save as new column

```{r}
Crime_clean$AddressCrime
Crime_clean$AddressCrime_new <- gsub("\\s+", " ", Crime_clean$AddressCrime)
Crime_clean$AddressCrime_new
```

# CriminalCharacteristicCodes contains multiple codes in one column separated by spaces; split into multiple columns

```{r}
Crime_clean$CriminalCharacteristicCodes_new =  Crime_clean$CriminalCharacteristicCodes

library(tidyverse)
cols <- paste0("Code_", 1:10)
Crime_clean2 <- Crime_clean %>% separate(CriminalCharacteristicCodes_new, sep = " ", into = cols )
View(Crime_clean2)
```

# Adding labels to gender variable

```{r}
Crime_clean2 <- Crime_clean2 %>%
  mutate(VictimSex = fct_recode(VictimSex, 
                                 "Male" = "M", 
                                 "Female" = "F", 
                                 "Unknown" = "X")) 
str(Crime_clean2)
```

# Adding labels to ethnicity variable

```{r}
Crime_clean2 <- Crime_clean2 %>%
  mutate(VictimEthnicity = fct_recode(VictimEthnicity,
                                 "Other Asian" = "A", 
                                 "Black" = "B", 
                                 "Chinese" = "C",
                                 "Filipino" = "F",
                                 "Hispanic/Latin/Mexican" = "H",
                                 "American Indian/Alaskan Native" = "I",
                                 "Japanese" = "J",
                                 "Korean" = "K",
                                 "Other" = "O",
                                 "Pacific Islander" = "P",
                                 "Samoan" = "S",
                                 "Vietnamese" = "V",
                                 "White" = "W",
                                 "Unknown" = "X")) 
str(Crime_clean2)
```

# Save as RDS

```{r}
folder = "DataCleanAndFormatted"

# Check if the folder exists
if (!dir.exists(folder)) {
  # Create the folder
  dir.create(folder)
  saveRDS(Crime_clean2,file.path(folder,"
                               Crime_formatted.RDS"))

} else {
  saveRDS(Crime_clean2,file.path(folder,"Crime_formatted.RDS"))

}
```